---
title: "Appendix 3: Creating a global measure of impairment"
output: word_document
---

```{r setup, include=FALSE}
#created 19th Jan 2018 by DVM Bishop

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(yarrr)
```

Our goal is to create a single scale reflecting global level of neurodevelopmental impairment. Data from initial parental telephone interview are available for all children. Data from language testing are available for all but two very low-functioning children, who were unable to attempt our tasks. Data from parental questionnaires (CCC-2 and SRS) and DAWBA DSM5 diagnoses are available for a subset. We use all available data for each child to create a scale by adding points as follows:

 History of speech problems = 1  
 Current help in mainstream school (support or special class or SLT) =1  
 Special school = 2  
 Dyslexia (test scores, unless no data , in which case report from parent interview) = 1  
 DLD (test scores, unless no data , in which case report from parent interview) = 1  
 ADHD (parental report or DAWBA diagnosis) = 1  
 Behaviour problems (DAWBA diagnosis of conduct disorder or clear description on interview) = 1  
 Autistic features: report from interview of definite diagnosis, or SRS = 90, or DAWBA diagnosis = 2  
 Low IQ (PIQ < 70 or refusal/inability to do battery - with exception of reading tests) = 1  
  
The R script to do the calculations is shown below.

```{r read in data}

readdir<-"/Users/dorothybishop/Dropbox/ERCAdvanced/Project SCT analysis/data from redcap/"
writedir<-"/Users/dorothybishop/Dropbox/ERCAdvanced/Project SCT analysis/SCT_ASD_analysis/Project_files/Data/"

#load redcap data 
sctdat<-read.csv(paste0(readdir,'SCTData_DATA_2018-01-19_1008.csv'))
sctdat<-filter(sctdat,sctdat$trisomy<9) #remove isochromosome case

twindat<-read.csv(paste0(readdir,'TwinsData_DATA_2018-01-19_1439.csv'))
twindat$dyslexia<-twindat$dld_rd%%10 #modulus 10 , ie last digit
twindat$lang_disorder<-round((twindat$dld_rd-4.5)/10) #first digit is dld code
twindat<-select(twindat,record_id,randomtwininclude,partial_testing,schooling,
                dyslexia,lang_disorder,slt,srs_t_score,
                neurodev_diagnosis,splang_conc,adhd_comb_dsm_r1,adhd_hyp_dsm_r1,adhd_inatt_dsm_r1,
                asd_dsm_r1,conduct_dsm_r1)
twindat$Group<-1
ntwin<-nrow(twindat)
sctdat<-select(sctdat,record_id,trisomy,partial_testing,schooling,
                dyslexia,lang_disorder,slt,srs_t_score,
                neurodev_diag,splang_conc,adhd_comb_dsm_r1,adhd_hyp_dsm_r1,adhd_inatt_dsm_r1,
                asd_dsm_r1,conduct_dsm_r1)
sctdat$Group<-0
colnames(sctdat)[2]<-'subgroup'
colnames(twindat)[2]<-'subgroup'
colnames(sctdat)[9]<-'neurodev_diagnosis'
nsct<-nrow(sctdat)
alldat<-rbind(sctdat,twindat)
alldat$Group<-as.factor(alldat$Group)
levels(alldat$Group)<-c('SCT','Comparison')
nrows<-nrow(alldat)


#note how many have missing data
a<-which(is.na(sctdat$adhd_comb_dsm_r1))
 b<-which(sctdat$srs_t_score>990)
 c<-which(is.na(twindat$adhd_comb_dsm_r1))
 d<-which(is.na(twindat$srs_t_score))

 print(paste('Total N twins = ',ntwin))
 print(paste('Twins with SRS = ',ntwin-length(d)))
 print(paste('Twins with DAWBA = ',ntwin-length(c)))

 print(paste('Total N SCT = ',nsct))
 print(paste('SCT with SRS = ',nsct-length(b)))
 print(paste('SCT with DAWBA = ',nsct-length(a)))

######################################################################################
# NB coding for slt is:
#0, never ; 1, preschool only; 2, beyond 4 yr; 3, ongoing; 8, assessed only; 9, no information

# Neurodev diagnosis codes:
# 0 none; others coded as all applicable from list:
#  1 ADHD, 2 APD, 3 ASD, 4 behav, 5 dyscalc, 6 dyslexia, 7 dyspraxia, 8  DLD/SLI/LD, 9 ID/GDD

#Lang dis from test scores; 0, no; 1, subclinical; 2, yes; 8, iq< 70; 9, no test results
#Dyslexia from test scors :0, no; 1, yes; 8, piq< 70; 9, no test results

#lang_concerns from parental report coded as splang_conc:
#0, never; 1, past; 2, continuing mild; 3, continuing severe; 9, unclear
#[for twins we also have code 4 if just concerns re reading; this is ignored here]

#SLT from parent report: 0, never; 1, preschool only; 2, beyond 4 yr; 3, ongoing; 8, assessed only; 9, no information

# School code
#1, mainstream no help; 2, mainstream with help; 3, special class/unit; 4, special school; 5, home schooled; 8, other; 9, dk

alldat$global_neurodev_rating<-0 #Initialise to zero
w<-which(is.na(alldat$srs_t_score)) #Recode NA to 999 for SRS
alldat$srs_t_score[w]<-999

temp<-alldat$global_neurodev_rating
w<-unique(which(alldat$slt==1),which(alldat$slt==8)) #Cases with preschool SLT or assessed by SLT
alldat$global_neurodev_rating[w]<-alldat$global_neurodev_rating[w]+1

#Now code so can add one point for help in mainstream or ongoing SLT or in language unit
w1<-c(which(alldat$schooling==2),which(alldat$schooling==3)) #help in mainstream/lang unit

alldat$global_neurodev_rating[w1]<-alldat$global_neurodev_rating[w1]+1

#add 2 points if attending special school
w<-which(alldat$schooling==4)
alldat$global_neurodev_rating[w]<-alldat$global_neurodev_rating[w]+1

#add 1 point if PIQ < 70 or not completed
w<-which(alldat$piq<70)
w1<-which(alldat$piq>996)
w2<-which(alldat$partial_testing>199) #failed to complete test battery (not because of age)
allw<-unique(w,w1,w2)
alldat$global_neurodev_rating[allw]<-alldat$global_neurodev_rating[allw]+1

## Next bits done in a loop because criteria not captured in a single code

for (i in 1:nrows){
  
  # add 1 to code if meets language test criteria for dyslexia OR (if no data) has diagnosis of this 
  # reported on parent interview
  wd<-NA
  temp<-alldat$dyslexia[i] #coding according to test battery, 1 if dyslexic
  if(temp==9){
    wd<-unlist(gregexpr(pattern ='6',toString(alldat$neurodev_diagnosis[i]))) #dyslexia code is 6
    #wd is one if 6 is included in neurodev_diag
  }
  if(length(wd)<1){temp=0}
  if (temp>0){alldat$global_neurodev_rating[i]<-alldat$global_neurodev_rating[i]+1}

#Add 1 to code if evidence of ADHD on parental interview or DAWBA
  w2<-unlist(gregexpr(pattern ='1',toString(alldat$neurodev_diagnosis[i]))) #ADHD code is 1
  if(w2==0){
  w2<-max(alldat$adhd_comb_dsm_r1[i],alldat$adhd_hyp_dsm_r1[i],alldat$adhd_inatt_dsm_r1[i])
  }
if (w2>0){alldat$global_neurodev_rating[i]<-alldat$global_neurodev_rating[i]+1}
  
# add 1 to code if meets language test criteria for lang_disorder OR (if no data) has diagnosis of this 
# reported on parent interview

temp<-alldat$lang_disorder[i] #coding according to test battery, 2 if with poor comp, 1 otherwise
w1<-temp
if(w1==2){w1<-1} #just one point added regardless of whether lang_dis code is 1 or 2
if(temp==9){ #no data on language tests so use parent interview
  w1<-unlist(gregexpr(pattern ='8',toString(alldat$neurodev_diagnosis[i]))) #DLD code is 8
  if(alldat$slt[i]==3){w1<-1} #regardless of diagnosis, ongoing SLT counts as DLD
  if(alldat$splang_conc[i]==3){w1<-1}#also serious language concerns count as DLD
} 
if (w1>0){alldat$global_neurodev_rating[i]<-alldat$global_neurodev_rating[i]+w1}
#NB more severe language problems with poor comprehension get addition of 2 points

# add 1 to code if significant behaviour problems on interview or DAWBA

  w1<-unlist(gregexpr(pattern ='4',toString(alldat$neurodev_diagnosis[i]))) #behav problems code is 4
  w2<-alldat$conduct_dsm_r1[i]
if (is.na(w2)){w2<-0}
  if (max(w1,w2)>0){alldat$global_neurodev_rating[i]<-alldat$global_neurodev_rating[i]+1}
  
# add 2 to code if ASD on interview or DAWBA or SRS is 90 or more

w1<-unlist(gregexpr(pattern ='3',toString(alldat$neurodev_diagnosis[i]))) #ASD code is 3
w2<-alldat$asd_dsm_r1[i]
if(is.na(w2)){w2<-0}
w3<-0

if(alldat$srs_t_score[i]>89) {w3<-1} #SRS t score 90

     if(alldat$srs_t_score[i]>900){w3<-0} 
if (max(w1,w2,w3)>0){alldat$global_neurodev_rating[i]<-alldat$global_neurodev_rating[i]+2}
#vertically jittered data created so all points visible on plot
alldat$global_jittered[i]<-alldat$global_neurodev_rating[i]+.5*runif(1)-.25
}
pirateplot(formula = alldat$global_jittered~ Group,
           point.o = .5,
           bar.f.o=.0,
           inf.f.o=.2,
           bean.b.o=.5,
           jitter=.1,
           data = alldat,
           ylab='Global rating',
           ylim=c(0,10),
           main="Distribution of global impairment\n in SCT and Comparison groups")


```
