---
title: "Coding_agreement_Di_Do"
author: "dorothy"
date: "29/11/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Coding of papers on CNTNAP2

 codes are: *C: cell
           *A: animal model
          *S: case study
            *M: group study of mutation/rare variant/CNV
           *V: group study of common variant
          * O: other (includes reviews/not relevant/etc)
           
Papers on CNTNAP2 from Scopus search were coded from information in Abstract.

Directory is "~/Dropbox/ERCadvanced/project SCT analysis/SCT genetic analysis/biblio_data/gene_specific_csv/

File is 'CNTNAP2_papers_done.csv': has columns for DB and DN


```{r agree,echo = FALSE}
readdir<-"~/Dropbox/ERCadvanced/project SCT analysis/SCT genetic analysis/biblio_data/gene_specific_csv/"
genefile<-paste0(readdir,'CNTNAP2_papers_done.csv')
mygenedata<-read.csv(genefile)
myagree<-table(mygenedata$Code,mygenedata$Di.code)
#Agreement based on all codes
perc_agree<-sum(diag(myagree))/sum(myagree)
kable(myagree,caption='Di rating codes along the top, Dorothy rating codes in rows')
paste0('Percentage agreement is ',round(100*perc_agree),'%')

#Recode into just V, M and other
w<-c(which(mygenedata$Di.code=='A'),which(mygenedata$Di.code=='C'),
     which(mygenedata$Di.code=='S'))
mygenedata$Di.shortcode<-mygenedata$Di.code
mygenedata$Di.shortcode[w]<-'O'
w1<-c(which(mygenedata$Code=='A'),which(mygenedata$Code=='C'),
     which(mygenedata$Code=='S'))
mygenedata$shortcode<-mygenedata$Code
mygenedata$shortcode[w1]<-'O'
myagreeshort<-table(mygenedata$shortcode,mygenedata$Di.shortcode)
myagreeshort<-myagreeshort[c(3,4,6),c(3,4,6)]
perc_agreeshort<-sum(diag(myagreeshort))/sum(myagreeshort)

#agreement for shortened codes
kable(myagreeshort,caption='3 category only: Di rating codes along the top, Dorothy rating codes in rows')
perc_agreeshort<-sum(diag(myagreeshort))/sum(myagreeshort)
paste0('Percentage agreement is ',round(100*perc_agreeshort),'%')
```


